Probabilities of polar aligned disks in binary star systems

### 2024-06-05 src/tex/ms.tex
Add more figures, get zenodo working
commit all scripts Some are just for the poster. Also finish the methods section

\newcommand{\RGM}[1]{\textcolor{cyan}{#1}}

@@

\section{Methods}
\label{sec:methods}
We use the N-body code {\sc rebound} \citep{rebound} to simulate a 3-body system. The code used to run these simulations is publically
available\footnote{\codeurl}.

\subsection{Simulation Setup}
\label{subsec:setup}

\begin{table*}
    \begin{tabular}{cccp{0.5\textwidth}}
        \hline
        name & symbol & type & description \\
        \hline
        \texttt{mass\_binary} & $M_b$ & \texttt{float} & The total mass of the binary in solar masses. \\
        \texttt{mass\_fraction} & $f_b$ & \texttt{float} & The ratio between the secondary mass and the binary mass $M_2/M_b$. \\
        \texttt{semimajor\_axis\_binary} & $a_b$ & \texttt{float} & The semimajor axis of the binary orbit. \\
        \texttt{eccentricity\_binary} & $e_b$ & \texttt{float} & The eccentricity of the binary orbit. \\
        \texttt{mass\_planet} & $m_p$ & \texttt{float} & The mass of the third body. \\
        \texttt{semimajor\_axis\_planet} & $a_p$ & \texttt{float} & The semimajor of the third body orbiting the binary center of mass. \\
        \texttt{inclination\_planet} & $i$ & \texttt{float} & The mutual inclination of the binary and the third body. \\
        \texttt{lon\_ascending\_node\_planet} & $\Omega$ & \texttt{float} & The longitude of the ascending node of the third body. \\
        \texttt{true\_anomaly\_planet} & $\nu$ & \texttt{float} & The true anomaly of the third body. \\
        \texttt{gr} & - & \texttt{bool} & Whether to include effects from general relativity. \\
        \texttt{eccentricity\_planet} & $e_p$ & \texttt{float} & The eccentricity of the third body's orbit. \\
        \texttt{arg\_pariapsis\_planet} & $\omega$ & \texttt{float} & The argument of pariapsis of the third body's orbit. \\
        \hline
    \end{tabular}
    \caption{Simulation parameters}
    \label{tab:params}
\end{table*}

We use {\sc rebound}'s Python interface to initialize our simulation. Table \ref{tab:params} describes the simulation's
input parameters. We integrate the system using the IAS15 15th order Gauss-Radau integrator \citep{reboundias15}
until we can determine the system's state.

\subsubsection{Determining the state}
\label{subsubsec:state}
We integrate the system, checking its state every 5 orbits of the third body.
We describe the type of orbit by computing the mutual inclination $i$ and the
longitude of the ascending node $\Omega$, which are defined:

\begin{equation}
    \cos{i} = \frac{\bm{L} \cdot \hat{\bm{z}}}{\|\bm{L}\|}
    \label{eq:inclination}
\end{equation}

@@

\begin{equation}
    \Omega = \arctantwo (\bm{L} \cdot \hat{\bm{x}}, -\bm{L} \cdot \hat{\bm{y}})
    \label{eq:omega}
\end{equation}

Where $\bm{L}$ is the angular momentum of the third body.
The coordinate system is defined relative to the binary, with the $z$-axis to be along the
angular momentum vector, the $x$-axis to be along the eccentricity vector, and the
$y$-axis so that the coordinate system is right-handed.

\begin{equation}
    \begin{aligned}
        \hat{\bm{z}} & = \bm{L}/\|\bm{L}\| \\
        \hat{\bm{x}} & = \bm{e}/\|\bm{e}\| \\
        \hat{\bm{y}} & = - \hat{\bm{x}} \times \hat{\bm{z}}
        \label{eq:coords}
    \end{aligned}
\end{equation}

The final state of the three-body system is assessed according to two criteria:

\begin{enumerate}
    \item Has $\Delta\Omega$ changed sign? In a precessing orbit $\Omega$ changes monotonically. If 
    $\frac{d\Omega}{dt}$ changes from positive to negative (or vise-versa) then we can be sure we have found a librating state.
    \label{item:omega}
    \item Has $i~\sin{\Omega}$ crossed zero? In a librating state this quantity will never change sign. 
    \label{item:sign}
\end{enumerate}

If neither criteria are true then we do not yet have enough information, and we continue the integration.
If criterion \ref{item:omega} is true then the system is librating. If criterion \ref{item:sign} is true then the system is precessing.
There is no configuration that would allow both criteria to be true.

Figure \ref{fig:sim_example} shows a set of example simulations along with their determined states.
These simulations all consist of a central $1 M_\odot$ binary of equal mass stars in an orbit with $a=0.2$ and $e=0.4$.
A massless third body is orbiting at a distance of $a=1$.
Visualizing the system in this way makes it clear why we have chosen our two criteria. Librating orbits will never cross the x-axis
and will cause the polar angle ($\Omega$) to turn back before reaching $0$ or $\pi$. Precessing orbits (both prograde and retrograde)
travel approximately in circles with near-constant $i$.

@@

[\textcolor{red}{Cherry Creek}]

>>>>

\newcommand{\TJ}[1]{\textcolor{orange}{#1}}
\newcommand{\RGM}[1]{\textcolor{cyan}{#1}}

@@

\section{Methods}
\label{sec:methods}
We describe two methods to approximate the evolution of a binary-disk system, each optimized for a different balance of computational cost and physical realism.
First, we use the N-body code {\sc REBOUND} \citep{rebound} to simulate the disk using a point mass. The second method employs a Runge-Kutta-Fehlburg integrator to solve
equations (7-10) from \citet{martin2019}. These equations treat the central binary in the quadrupole approximation, but can be solved extremely efficiently.
The code used to run these simulations is publically available\footnote{\codeurl}.

\subsection{Geometric considerations}
\label{subsec:setup}

The dynamical state of the disk (i.e. prograde circulation, libration, etc.) is determined by tracking two relevant orbital elements:
mutual inclination $i$ and longitude of ascending node $\Omega$. Each simulation method gives as an output the direction of the disk angular momentum relative to
the binary $\bm{l}$. We define a coordinate system relative to the binary, with the $z$-axis to be along the
angular momentum vector, the $x$-axis to be along the eccentricity vector, and the
$y$-axis so that the coordinate system is right-handed.

The relevant coordinates are then defined

\begin{equation}
    \cos{i} = \frac{\bm{l} \cdot \hat{\bm{z}}}{\|\bm l\|}
    \label{eq:inclination}
\end{equation}

@@

\begin{equation}
    \Omega = \arctantwo (\bm{l} \cdot \hat{\bm{x}}, -\bm{l} \cdot \hat{\bm{y}})
    \label{eq:omega}
\end{equation}

Given some initial $(i,\Omega)$, the system can be evolved forward in time to recover the familiar diagram shown in Figure \ref{fig:sim_example}.
\TJ{cite papers that have similar figures.}
This visualiztaion also allows the dynamical state to be resolved programically, as a librating state will repeatedly cross the $y$-axis and never
cross the $x$-axis, and a circulating state will sequentially cross each axis in an order that can determine if it is prograde or retrograde.

\subsection{Using {\sc REBOUND}}
\label{subsec:reb}

\begin{table*}
    \begin{tabular}{cccp{0.5\textwidth}}
        \hline
        name & symbol & type & description \\
        \hline
        \texttt{mass\_binary} & $M_b$ & \texttt{float} & The total mass of the binary in solar masses. \\
        \texttt{mass\_fraction} & $f_b$ & \texttt{float} & The ratio between the secondary mass and the binary mass $M_2/M_b$. \\
        \texttt{semimajor\_axis\_binary} & $a_b$ & \texttt{float} & The semimajor axis of the binary orbit. \\
        \texttt{eccentricity\_binary} & $e_b$ & \texttt{float} & The eccentricity of the binary orbit. \\
        \texttt{mass\_planet} & $m_p$ & \texttt{float} & The mass of the third body. \\
        \texttt{semimajor\_axis\_planet} & $a_p$ & \texttt{float} & The semimajor of the third body orbiting the binary center of mass. \\
        \texttt{inclination\_planet} & $i$ & \texttt{float} & The mutual inclination of the binary and the third body. \\
        \texttt{lon\_ascending\_node\_planet} & $\Omega$ & \texttt{float} & The longitude of the ascending node of the third body. \\
        \texttt{true\_anomaly\_planet} & $\nu$ & \texttt{float} & The true anomaly of the third body. \\
        \texttt{gr} & - & \texttt{bool} & Whether to include effects from general relativity. \\
        \texttt{eccentricity\_planet} & $e_p$ & \texttt{float} & The eccentricity of the third body's orbit. \\
        \texttt{arg\_pariapsis\_planet} & $\omega$ & \texttt{float} & The argument of pariapsis of the third body's orbit. \\
        \hline
    \end{tabular}
    \caption{ {\sc REBOUND} Simulation parameters}
    \label{tab:params}
\end{table*}

We use {\sc REBOUND}'s Python interface to initialize our simulation. Table \ref{tab:params} describes the simulation's
input parameters. We integrate the system using the IAS15 15th order Gauss-Radau integrator \citep{reboundias15}
until we can determine the system's state.

Not much can be learned from a single simulation as it only allows us to map from one point in $(i,\Omega)$ space to an end state. To infer the behavior of a population
(and from that compute a polar fraction) we employ a Monte Carlo (MC) method, sampling isotropically on the $(i,\Omega)$ sphere. That is, $\Omega \in \mathcal{U}(0,2\pi)$
and $i \in \{\arcsin{(2u-1)} + \frac{\pi}{2} \mid u \in \mathcal{U}(0,1)\}$.

\begin{figure*}
    \begin{centering}
        \includegraphics{figures/mc_example.pdf}
        \caption{Monte Carlo results for a $1 M_\odot$ equal mass binary with $e_b=0.4$ for $j=0,0.5,1$. The position of each
        point shows the initial condition of the simulation, and the color denotes the dynamical state. Note that some are marked
        as ``unknown''. This is because we set a 1000 orbit limit on the integration for the sake of computational time.}
        \script{mc_example.py}
        \label{fig:mc}
    \end{centering}
\end{figure*}

Each MC run is initialized with the conditions of the binary as well as the angular momentum of the disk. We then sample $i$ and $\Omega$ in batches of four simulations,
using a bootstrap metric after each batch to compute the polar fraction. Sampling stops when the bootstrap-computed confidence interval reaches a prescribed value. A 
typical simulation takes on the order of $0.1-1$ s, so we save the results of each simulation in an SQLite database between batches. When an MC sampler is initialized
it queries this database for all relevant passed simulations, ingests them, and only runs new simulations if a more stringent confidence interval is requested.

\subsection{Runge-Kutta-Fehlburg method}
\label{subsec:rk}

In addition to using {\sc REBOUND} in Section \ref{subsec:reb} we can solve the same problem using a slightly more analytical approach. \citet{martin2019} found
a set of first-order differential equations (their Equations 7-10) based on previous work by \citet{farago2010} that describes the coupled behavior of the
disk angular momentum direction and the binary eccentricity. We solve these equations using the Runge-Kutta-Fehlberg (RKF) method \TJ{ref?} -- a numerical method similar
to the common 4th order Runge-Kutta integrator, but that computes a 5th order solution to allow for a variable step size. The cost of this method is that it treats
the binary in the quadrupole approximation, meaning we are not sensitive to the effects of the binary mass fraction, but the computational cost is extremely low; we are
able to run $10^4$-$10^5$ simulations per second -- about 10,000 times faster than {\sc REBOUND}.

The added computational efficiency means this method is not restricted to an MC method with hundreds of samples; we can compute a grid of $\sim 10^6$ in
less than a minute to get a very accurate measure of the polar fraction.

\begin{figure}
    \begin{centering}
        \includegraphics[options]{figures/compare_massive.pdf}
        \caption{
            Contour lines denoting solution boundaries found via the RKF method compared to MC results of the same conditions. Binary setup is the same
            as in Figure \ref{fig:mc}, but for $j=0.66$, which is its critical value \citep[see][]{martin2019,abod2022}. The grid here includes 250,000 points.
            There is some disagreement between RKF and {\sc REBOUND}, which we attribute to misclassification due to behavior on the orbital timescale (i.e. 
            small variations in $\Omega$ that cause the simulation to temporarily cross the $y$-axis twice, giving the false appearance of libration). \TJ{We can fix this}
        }
        \script{compare_massive.py}
        \label{fig:rkf}
    \end{centering}
\end{figure}

@@
